---
title: "Overview and some plots of the data"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged
---

Getting an overview of the data

## 0. Libraries
```{r}
library(tidyverse)
library(readxl)
library(pander)
```

## 1. Folders and files    
Content of folders (see code)  
```{r}
dir("Datasett")
dir("Datasett/River data (from OKA)")
dir("Datasett/hydrografi")
dir("Datasett/Bløtbunn")
dir("Datasett/hardbunn_kopi")
dir("Datasett/Plankton")
```

## 2. River data   
Content of folders (see code)
```{r}
dir("Datasett/River data (from OKA)/Annual mean flows")
dir("Datasett/River data (from OKA)/Monthly loads")
dir("Datasett/River data (from OKA)/Concentrations (individual samples)")
dir("Datasett/River data (from OKA)/Monthly flow-weighted concentrations")
```

### a. Data; example: Storelva loads
```{r}
df <- read_excel("Datasett/River data (from OKA)/Monthly loads/Storelva_monthly loads.xlsx")  
# df <- read_excel("Datasett/River data (from OKA)/Monthly loads/Storelva_monthly loads.xlsx", skip = 1)  
# summary(df)

df <- as.data.frame(df)[-1,]
# colnames(df) %>% dput()
vars <- c("TrspTot TOTN", "TrspTot NO3-N", "TrspTot NH4-N", "TrspTot TOTP", 
          "TrspTot TOC", "TrspTot ALK", "TrspTot Ca", "DisTot")
for (var in vars)
  df[,var] <- as.numeric(df[,var])
df$Time <- with(df, lubridate::ymd(paste(Year, Month, "15")))

# Add "_" in column names (TrspTot Ca -> TrspTot_Ca)
colnames(df) <- sub(" ", "_", colnames(df), fixed = TRUE)

tb <- df %>% 
  gather("Variable", Value, TrspTot_TOTN:DisTot) %>%
  filter(!is.na(Value)) %>%
  xtabs(~Year + Variable, .)
pandoc.table(tb, style = "rmarkdown")
```

### b. Plot Storelva loads  
```{r}
gg <- df %>%
  gather("Variable", Value,  TrspTot_TOTN:DisTot) %>%
  ggplot(aes(Time, Value)) + 
    geom_line() +
    facet_wrap(~Variable, scales = "free_y")
gg
```

### c. Plot annual flow  
```{r}
df <- read_excel("Datasett/River data (from OKA)/Annual mean flows/Mean annual flow.xlsx")
# df <- read_excel("Datasett/River data (from OKA)/Monthly loads/Storelva_monthly loads.xlsx", skip = 1)  
# summary(df)

df <- as.data.frame(df)[-1,]
colnames(df) <- sub(" ", "_", colnames(df), fixed = TRUE)
# colnames(df) %>% dput()
vars <- c("DisMean")
for (var in vars)
  df[,var] <- as.numeric(df[,var])

ggplot(df, aes(Year, DisMean)) + geom_line() + facet_wrap(~Station_name)

```

## 3. Hydrografi  
### a. Data  
Tables of coverages of depth and seasons  
```{r}
load("Datasett/Hydrografi/Arendal_allvars_1990_2016.Rdata")
Df.Arendal$Month <- Df.Arendal$Dato %>% as.character() %>% substr(6,7) %>% as.numeric()
Df.Arendal$Year <- Df.Arendal$Dato %>% as.character() %>% substr(1,4) %>% as.numeric()

# summary(Df.Arendal)
xtabs(~Depth + Month, Df.Arendal)
xtabs(~Year + Month, Df.Arendal)
```

### b. Plot salinity, depth 10 m
```{r}
ggplot(Df.Arendal %>% filter(Depth == 10), aes(Dato, Salt)) +
  geom_line() + geom_point()
```

### c. Salinity, time series by month and depth
```{r}
Df.Arendal %>%
  group_by(Year, Month, Depth) %>%
  summarise_all(mean) %>%
  ggplot(aes(Year, Salt)) + 
    geom_line() + 
    facet_grid(Month~Depth)
```


## 4. Soft bottom fauna  
### a. Data
```{r}
df_blot_b35 <- read_excel("Datasett/Bløtbunn/Klimaoverblikk bløtbunn_data til Helene og Dag.xlsx", sheet = "B35_artsliste")
colnames(df_blot_b35)[1] <- "Species"

df_blot_b05 <- read_excel("Datasett/Bløtbunn/Klimaoverblikk bløtbunn_data til Helene og Dag.xlsx", sheet = "B05_artsliste")
colnames(df_blot_b05)[1] <- "Species"

df_blot_ind <- read_excel("Datasett/Bløtbunn/Klimaoverblikk bløtbunn_data til Helene og Dag.xlsx", sheet = "indekser_sedimentparametere")

cat("Number of species:", nrow(df_blot_b35), "\nNumber of years:", ncol(df_blot_b35), "\n")
```

### b. Station B35, boxplot of the 10% most abundant species   
```{r}
df <- df_blot_b35 %>%
  gather("Year", "N", -Species) %>%
  mutate(N = ifelse(is.na(N),0,N)) %>%
  group_by(Species) %>%
  mutate(Median_per_species = median(N))

cat("Percentiles of median abundance per species:\n")
quantile(df$Median_per_species, (0:10)/10, na.rm = TRUE)

df %>%
  filter(Median_per_species >= 2) %>%
  ggplot(aes(Species, N)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
 
## 5. Hard-bottom fauna  
### a. Data (already chewed a bit on)
```{r}
# dir("Datasett/hardbunn_kopi")
# readLines("Datasett/hardbunn_kopi/HBanalysesett.csv", 2)
df <- readr::read_csv2("Datasett/hardbunn_kopi/HBanalysesett.csv")
```

### b. Plot DCA
```{r}
df %>%
  gather("DCA_axis", "Value", DCA1:DCA4) %>%
  ggplot(aes(Year, Value)) +
    geom_line() +
    facet_grid(Site~DCA_axis)
```

### c. Plot NMDS  
```{r}
df %>%
  gather("NMDS_axis", "Value", NMDS1:NMDS2) %>%
  ggplot(aes(Year, Value)) +
    geom_line() +
    facet_grid(Site~NMDS_axis)
```

### d. Plot DCAstrand
```{r}
df %>%
  gather("DCA_axis", "Value", DCA1strand:DCA4strand) %>%
  ggplot(aes(Year, Value)) +
    geom_line() +
    facet_grid(Site~DCA_axis)
```

### e. Plot species
```{r, fig.width=9}
df %>%
  gather("Species", "Value", `LGD.Halidrys siliquosa`:`LGD.Rhodomela confervoides`) %>%
  mutate(Species = sub("LDG.", "", Species)) %>%
  ggplot(aes(Year, Value)) +
    geom_line() +
    facet_grid(Site~Species)
```

## 6. Phytoplankton
### a. Data   
* Note: deleted the empty second row manually, and one empty column 
* Also copied the data (just the range with data, A1:V471) to a new sheet
```{r}
df <- read_excel("Datasett/Plankton/Planteplankton Arendal.xlsx") # range = "A1:V471"
df$Year <- lubridate::year(df$Dato)
df$Month <- lubridate::month(df$Dato)
xtabs(~Year + Month, df)
```

### b. Plot time series of main groups
```{r, fig.height=3.5}
df %>%
  gather("Group", "Value", Kiselalger:Flagellater) %>%
  ggplot(aes(Dato, Value)) +
    geom_line() +
    facet_wrap(~Group, scale = "free_y")
```

### c. Plot time series of main groups by month
```{r}
df %>%
  gather("Group", "Value", Kiselalger:Flagellater) %>%
  group_by(Year, Month, Group) %>%
  summarize(Value = mean(Value)) %>%
  filter(Month %in% c(2,4,6,8,10)) %>%
  ggplot(aes(Year, Value)) +
    geom_line() +
    facet_grid(Group~Month, scale = "free_y")
```

### d. Plot time series of 6 main taxa (by median abundance) by month
```{r, fig.height=6}
mean_abundance <- apply(df %>% select(Cyanophycea:Chlorophycea), 2, mean, na.rm = TRUE) %>% sort() %>% rev()
median_abundance <- apply(df %>% select(Cyanophycea:Chlorophycea), 2, median, na.rm = TRUE) %>% sort() %>% rev()
taxa <- names(median_abundance) %>% head(6)
df %>%
  gather("Taxon", "Value", Cyanophycea:Chlorophycea) %>%
  group_by(Year, Month, Taxon) %>%
  summarize(Value = mean(Value)) %>%
  filter(Month %in% c(2,4,6,8,10) & Taxon %in% taxa) %>%
  ggplot(aes(Year, Value)) +
    geom_line() +
    facet_grid(Taxon~Month, scale = "free_y")
```

### e. Plot time series of 5 lesser taxa (by median abundance) by month
```{r, fig.height=6}
taxa <- names(median_abundance) %>% tail(5)
df %>%
  gather("Taxon", "Value", Cyanophycea:Chlorophycea) %>%
  group_by(Year, Month, Taxon) %>%
  summarize(Value = mean(Value)) %>%
  filter(Month %in% c(2,4,6,8,10) & Taxon %in% taxa) %>%
  ggplot(aes(Year, Value)) +
    geom_line() +
    facet_grid(Taxon~Month, scale = "free_y")
```


