---
title: "Organising soft-bottom data for PCA"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged    
---


## 0. Libraries
```{r}
library(tidyverse)
library(readxl)
# install.packages("vegan")
library(vegan)
```

### Function for finding text in R files  
```{r}

search_text <- function(string, dir = ".", extension = c("R","Rmd"), pattern = NULL, fixed = FALSE, ignore.case = TRUE, deep = FALSE){
  if (!is.null(extension) & !is.null(pattern))
    warning("Pattern will take predence over extension")
  if (is.null(pattern)) {
    ext <- paste(extension, collapse = "|")
    pattern <- paste0("\\.(", paste(extension, collapse = '|'), ")$")
	}
  fn <- list.files(path = dir, pattern = pattern, full.names = TRUE, recursive = deep)
  search <- plyr::alply(fn, 1, function(x) grep(string,  readLines(x, warn = FALSE), fixed = fixed, ignore.case = ignore.case), .progress = "text")
  fn[plyr::laply(search, length) > 0]	
  }
```

## 1. Check organisation/analysis of hard-bottom data from Guri's scripts   
### a. Content of folders (see code)  
```{r}
dir("Datasett/hardbunn_kopi")
dir("Datasett/hardbunn_kopi/r workspace")
#
# Used this to explore where different data sets occur:
# search_text("ord1.df", "Datasett/hardbunn_kopi/r workspace")
# 
```

### b. Read 'transekt.df' which Guri used for ordination  
Does not include ord1.df, which is used in 'Klimaoverblikk_hardbunn.Rmd', but does include a lot of other stuff...  
Order of Guri's scripts:  
1) HBdata.R: HBdata -> Hbuse  
2) HBOrdinasjon.R: Hbuse -> HBagg -> transekt.df -> SiteSpec -> spec.m -> ordinasjon -> ord1.df  
```{r}
load("Datasett/hardbunn_kopi/r workspace/.RData")

# Let us delete everything except transekt.df
obj <- ls()
rm(list = obj[!obj %in% c("transekt.df")])

str(transekt.df)
```

### c. Create data for ordination ('spec.m') and perform ordination   
From Guri's code in 'HBOrdinasjon.R', except the first part     
* 'transekt.df' is in 'long format'  
* Is reshaped to broad format ('SiteSpec'). *Note* that this one is later combined with PCA output to create 'ord1.df'   
* For the ordination, the year and site columns are removed and NA -> 0 ('spec.m')   
```{r, warning=FALSE}
# table(transekt.df$Sitename)
transekt.df <- transekt.df %>%
  mutate(Site = substr(Sitename, 1,3), Year = as.numeric(substr(Sitename, 5,8))) %>%
  select(Site, Year, Species, Value)
# names(transekt.df) = c("Site", "Year", "Species", "Value")

# gjor om til vidt format
SiteSpec = reshape(transekt.df, idvar = c("Site", "Year"), timevar = "Species", direction = "wide")
# names(SiteSpec)

# Endrer navnene
artsnavn = names(SiteSpec)[-c(1:2)]
artsnavn = gsub("Value.", "", artsnavn)
artsnavn = gsub(" \n", "", artsnavn)
# artsnavn

# artsmatrise
spec.m = SiteSpec[, -c(1:2)]
names(spec.m) = artsnavn

# må erstatte NA med 0
spec.m[is.na(spec.m)] = 0

cat("Dimensions of 'spec.m':\n", dim(spec.m))
```

### d. Perform ordination
```{r}
# Transekt
ord1 <- decorana(spec.m)

# Pretty longish output
# summary(ord1) # viser arts- og rutescorer

# Short output
summary(ord1, display = "none")
```
### e. Test plot 1
```{r}
plot(ord1) # sorte sirkler er sites, røde pluss er arter
```

### f. Test plot 2
```{r}
plot(ord1, display = "sites", type = "n")
points(ord1, display = "sites", pch = 21, col = "red", bg = "yellow")
```

### g. Make data set for the plots done in 'Klimaoverblikk_hardbunn.Rmd'  
```{r}
# trekker ut akseskorer så det kan kobles til stasjon og år... i.e. Sitename
ord1.sites <- data.frame(scores(ord1, display = "sites"))

ord1.df <- data.frame(SiteSpec[,c(1:2)], ord1.sites)

str(ord1.df)
```

### h. Redone plots done in 'Klimaoverblikk_hardbunn.Rmd'   
* *BUT* they look a bit different... so the input data are probably not identical  
```{r}
# Innledende DCA - litt usikker på utfallet  
DCAplot.1 <- ggplot(ord1.df, aes(DCA1, DCA2, color = as.numeric(as.character(Year)))) + 
  geom_point() + 
  facet_wrap(~ Site, nrow=2) +
  labs(color = "År")

DCAplot.2 <- ggplot(ord1.df, aes(DCA3, DCA4, color = as.numeric(as.character(Year)))) + 
  geom_point() + 
  facet_wrap(~ Site, nrow=2) +
  labs(color = "År")

DCAplot.1
```


## 2. Reorganise soft-bottom fauna data and test ordination
### a. Data
```{r}
df_blot_b35 <- read_excel("Datasett/Bløtbunn/Klimaoverblikk bløtbunn_data til Helene og Dag.xlsx", sheet = "B35_artsliste")
colnames(df_blot_b35)[1] <- "Species"

df_blot_b05 <- read_excel("Datasett/Bløtbunn/Klimaoverblikk bløtbunn_data til Helene og Dag.xlsx", sheet = "B05_artsliste")
colnames(df_blot_b05)[1] <- "Species"

df_blot_ind <- read_excel("Datasett/Bløtbunn/Klimaoverblikk bløtbunn_data til Helene og Dag.xlsx", sheet = "indekser_sedimentparametere")

cat("b35, number of species:", nrow(df_blot_b35), ", Number of years:", ncol(df_blot_b35), "\n")
cat("b05, number of species:", nrow(df_blot_b05), ", Number of years:", ncol(df_blot_b05), "\n")
```

### b. Put on data long format, combine, and extract Site and Year separately
```{r}
df_long_1 <- df_blot_b05 %>%
  gather("Siteyear", "Value", B05_1990:B05_2016)
df_long_2 <- df_blot_b35 %>%
  gather("Siteyear", "Value", B35_1990:B35_2016)

df_long <- bind_rows(df_long_1, df_long_2) %>%
  mutate(Site = substr(Siteyear, 1,3), Year = as.numeric(substr(Siteyear, 5,8))) %>%
  select(Site, Year, Species, Value)

str(df_long)
```

### c. Reshape for ordination  
Reshape to wide format the tidyr way (instead of using reshape()) - gives us nice column names right away  
```{r}
SiteSpec <- df_long %>%
  spread(Species, Value)

# artsmatrise
spec.m = SiteSpec[, -c(1:2)]

# må erstatte NA med 0
spec.m[is.na(spec.m)] = 0

cat("Dimensions of 'spec.m':\n", dim(spec.m))
```

### d. Perform ordination
```{r}
# Transekt
ord1 <- decorana(spec.m)

# Pretty longish output
# summary(ord1) # viser arts- og rutescorer

# Short output
summary(ord1, display = "none")
```
### e. Test plot 1
```{r}
plot(ord1) # sorte sirkler er sites, røde pluss er arter
```

### f. Test plot 2
```{r}
plot(ord1, display = "sites", type = "n")
points(ord1, display = "sites", pch = 21, col = "red", bg = "yellow")
```

### g. Make data set for ggplot
```{r}
# trekker ut akseskorer så det kan kobles til stasjon og år... i.e. Sitename
ord1.sites <- data.frame(scores(ord1, display = "sites"))

ord1.df <- data.frame(SiteSpec[,c(1:2)], ord1.sites)

str(ord1.df)
```

### h. Redone plots similar to 'Klimaoverblikk_hardbunn.Rmd'   
```{r}
# Innledende DCA - litt usikker på utfallet  
DCAplot.1 <- ggplot(ord1.df, aes(DCA1, DCA2, color = as.numeric(as.character(Year)))) + 
  geom_point() + 
  facet_wrap(~ Site, nrow=2) +
  labs(color = "År")

DCAplot.2 <- ggplot(ord1.df, aes(DCA3, DCA4, color = as.numeric(as.character(Year)))) + 
  geom_point() + 
  facet_wrap(~ Site, nrow=2) +
  labs(color = "År")

DCAplot.1
```
