---
title: "Extract spring bloom phytoplankton data"
output: html_document
---


```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(lubridate)   # yday() for day-of-year
```


## 1. Data
### a. Read hydrography data
```{r}
load("Datasett/Hydrografi/Arendal_allvars_1990_2016.Rdata")
Df.Arendal$Month <- Df.Arendal$Dato %>% as.character() %>% substr(6,7) %>% as.numeric()
Df.Arendal$Year <- Df.Arendal$Dato %>% as.character() %>% substr(1,4) %>% as.numeric()

Df.Arendal$Date <- ymd_hms(paste(Df.Arendal$Dato, "00:00:00"))   # R's date format
```

### b. Read plankton data
```{r}
df_plank <- read_excel("Datasett/Plankton/Planteplankton Arendal.xlsx") # range = "A1:V471"
df_plank$Year <- lubridate::year(df_plank$Dato)
df_plank$Month <- lubridate::month(df_plank$Dato)
```

### c. Plankton: Select by depth  
0-30 or 5 m  
```{r}
xtabs(~Dyp, df_plank)

# Select
sel <- df_plank$Dyp %in% c("0-30 m", "5 m", "5m"); 
df_plank <- df_plank[sel,]

# Stats
cat("Select", sum(sel), "lines\n")
cat(mean(sel)*100, "% of the data")

```

## 2. Plot ChlA
### a. All year
```{r}
Df.Arendal %>%
  filter(Depth %in% 5) %>%
  mutate(Decade = paste0(10*floor(Year/10), "s")) %>%
  ggplot(aes(yday(Date), Klorofyll, group = Year, color = factor(Year))) + 
    geom_line() +
    facet_wrap(~Decade)
```

### b. Spring (January - May)
```{r}
Df.Arendal %>%
  filter(Depth %in% 5 & Month %in% 1:5) %>%
  mutate(Decade = paste0(10*floor(Year/10), "s")) %>%
  ggplot(aes(yday(Date), Klorofyll, group = Year, color = factor(Year))) + 
    geom_line() +
    facet_grid(Decade~.)
```
### c. Spring (January - May) log-scale
```{r}
yday_1feb <- yday(ymd(20010201))
yday_30apr <- yday(ymd(20010430))
Df.Arendal %>%
  filter(Depth %in% 5 & Month %in% 1:5) %>%
  mutate(Decade = paste0(10*floor(Year/10), "s")) %>%
  ggplot(aes(yday(Date), Klorofyll, group = Year, color = factor(Year))) + 
    geom_line() +
    geom_vline(xintercept = c(yday_1feb, yday_30apr)) +
    scale_y_log10() +
    facet_grid(Decade~.)
```

 
### d. Max chlorophyll February-April and February-May
```{r}
df <- Df.Arendal %>%
  filter(Depth %in% 5) %>% 
  group_by(Year) %>%
  summarize(FebApr = max(Klorofyll[Month %in% 2:4], na.rm = TRUE),
            FebMay = max(Klorofyll[Month %in% 2:5], na.rm = TRUE))
df %>%
  gather("Period", "Chla_max", FebApr, FebMay) %>%
  ggplot(aes(Year, Chla_max, color = Period)) + 
  geom_line()
cat("Lowest max value Feb-April:", min(df$FebApr), "\n")
```

## 3. Get bloom periods to use  
### a. Bloomperiod_01: Chl a over absolute threshold  
```{r}
threshold <- 1.1
Bloomperiod_01 <- Df.Arendal %>%
  filter(Depth %in% 5 & Month %in% 2:4 & Klorofyll >= threshold & !is.na(Klorofyll)) %>% 
  group_by(Year) %>%
  summarize(Date1 = min(Date), Date2 = max(Date))
```

### b. Plot periods
```{r}
df <- Df.Arendal %>%
  filter(Depth %in% 5 & Month %in% 1:5) %>%
  mutate(Decade = paste0(10*floor(Year/10), "s")) %>%
  mutate(Bloomperiod = Month %in% 2:4 & Klorofyll >= threshold & !is.na(Klorofyll))

ggplot(df, aes(yday(Date), Klorofyll, group = Year, color = factor(Year))) + 
    geom_line() +
    geom_point(data = df %>% filter(Bloomperiod), size = rel(3)) +
    geom_vline(xintercept = c(yday_1feb, yday_30apr)) +
    scale_y_log10() +
    facet_grid(Decade~.)

ggplot(df, aes(yday(Date), Klorofyll, group = Year, color = factor(Year), size = Bloomperiod)) + 
    geom_line() +
    scale_size_manual(values = c(0.5, 1.5)) +
    geom_point(data = df %>% filter(Bloomperiod), size = rel(3)) +
    geom_vline(xintercept = c(yday_1feb, yday_30apr)) +
    scale_y_log10() +
    facet_grid(Decade~.)
```

### c. Bloomperiod_02: Max Chl a
```{r}
Bloomperiod_02 <- Df.Arendal %>%
  filter(Depth %in% 5 & Month %in% 2:4 & Klorofyll >= threshold & !is.na(Klorofyll)) %>% 
  group_by(Year) %>%
  summarize(Date = Date[Klorofyll %in% max(Klorofyll)][1])
```

## 4. Pick data in plankton data set
### a. Create Bloom01 and BLoom2 variables (false/true) in plankton data set  
We make a "buffer zone" of one extra day in each direction (in 1998, max Chla is 6 April by plankton data are from 7 april)  
```{r}
df_plank <- df_plank %>% as.data.frame()
Bloomperiod_01 <- Bloomperiod_01 %>% as.data.frame()
Bloomperiod_02 <- Bloomperiod_02  %>% as.data.frame()
df_plank$Bloom01 <- FALSE
df_plank$Bloom02 <- FALSE
for (yr in unique(Bloomperiod_01$Year)){
  # yr <- 1998
  df <- Bloomperiod_01 %>% filter(Year %in% yr) 
  sel <- df_plank$Year %in% yr & df_plank$Dato >= (df$Date1[1] - days(1)) & df_plank$Dato <= (df$Date2[1] + days(1))
  df_plank$Bloom01[sel] <- TRUE
  df <- Bloomperiod_02 %>% filter(Year %in% yr) 
  sel <- df_plank$Year %in% yr & df_plank$Dato >= (df$Date[1] - days(1)) & df_plank$Dato <= (df$Date[1] + days(1)) 
  df_plank$Bloom02[sel] <- TRUE
}
xtabs(~Year + Bloom01, df_plank)
xtabs(~Year + Bloom02, df_plank)
df_plank %>% filter(Bloom02) %>% View()

```

### b. Extract plankton data - mean/median values based on Bloom01
```{r}
df_plank_bloom01_mean <- df_plank %>%
  filter(Bloom01) %>%
  group_by(Year) %>%
  summarize_at(vars(Cyanophycea:Flagellater), mean, na.rm = TRUE)

df_plank_bloom01_median <- df_plank %>%
  filter(Bloom01) %>%
  group_by(Year) %>%
  summarize_at(vars(Cyanophycea:Flagellater), median, na.rm = TRUE)
    
```

### b. Extract plankton data - values based on Bloom02, i.e. max ChlA
```{r}
df_plank_bloom02 <- df_plank %>%
  filter(Bloom02)

```

```{r}
write.csv(df_plank_bloom01_mean, "Data_produced/df_plank_bloom01_mean.csv", row.names = FALSE, quote = FALSE)
write.csv(df_plank_bloom01_median, "Data_produced/df_plank_bloom01_median.csv", row.names = FALSE, quote = FALSE)
write.csv(df_plank_bloom02, "Data_produced/df_plank_bloom02.csv", row.names = FALSE, quote = FALSE)
```



