---
title: "07 Plots of annual data"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged   
---

Mass production of time series plots including Theil-Sen regression line  
* Uses library 'openair' for plots    
  
Plots are saved in folder [Figures_07](Figures_07)  

Note: plotting uses a file [07_Plotdata.csv](07_Plotdata.csv) (manually made in Excel) to add y labels; this file can be extedned to also included headers to be added to each plot  

## 0. Libraries
```{r}
library(tidyverse)
library(readxl)
library(broom)
library(lubridate)
install.packages("openair")
library(openair)  

# library(pander)
```

## 1. Data  
```{r}
dat_a <- read.csv("Data_produced/06_dat_a.csv")
dat_q <- read.csv("Data_produced/06_dat_q.csv")

```

## 2. Test plot  
```{r}
varname <- "Dinoflagellater_med"
varname <- "Tetthet_Deep"
df <- dat_a %>% filter(Variable %in% varname)
df$date <- ymd(paste(df$Year, "07 01"))

# test plot
TheilSen(df, pollutant = "Value", main = varname, ylab = "Median during year",
         shade = "transparent", pch = 21, cex = 1.2, data.col = "black", fill = "firebrick")

```


## 3. Define function for making and saving plots   
```{r}
# write.table(tibble(unique(dat_a$Variable)), "clipboard", row.names = FALSE, quote = FALSE)
# plotdata <- read_xls("07_Plotdata.xlsx", sheet = 1)
plotdata <- read.csv2("07_Plotdata.csv", stringsAsFactors = FALSE)

TheilSen_save <- function(varname, ylabel){
  df <- dat_a %>% filter(Variable %in% varname)
  df$date <- ymd(paste(df$Year, "07 01"))
  png(sprintf("Figures_07/Annual_timeseries_%s.png", varname), width = 15, height = 10, units = 'cm',  
    res = 300, type="cairo", antialias="default")
  TheilSen(df, pollutant = "Value", main = varname, ylab = ylabel,
   shade = "transparent", pch = 21, cex = 1.2, data.col = "black", fill = "firebrick")
  dev.off()
}

# TheilSen_save("Flagellater_med", "Density")
# i <- 73
# i <- 1
# TheilSen_save(plotdata[i,"Variable"], plotdata[i, "Ylabel"])

```

## 4. Actually produce plots (takes perhaps 10-15 minutes)  
Set 'produce_plots <- TRUE' in order to actually produce plots
```{r}
produce_plots <- FALSE

if (produce_plots){
  for (i in 1:nrow(plotdata)){
    TheilSen_save(plotdata[i,"Variable"], plotdata[i, "Ylabel"])
    Sys.sleep(2)
  }
}


```




