---
title: "08 Regression models"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged   
---


Regression analysis of hydrographic and plankton variables.

Check out package dynlm for including lagged variables - see https://stackoverflow.com/a/13096824/1734247 

## 0. Libraries
```{r}
library(tidyverse)
library(readxl)
library(MASS)
library (ggplot2)
#install.packages("visreg")
library(visreg)
library(vegan)
library(lme4) # for mixed effect models
#install.packages("standardize")
library(standardize) # standardize data for modelling
#install.packages("afex")
library(afex) # adding p-values to lmer output

# library(pander)
```

## 1. Data  
### a. Read
```{r}
dat_a <- read.csv("Data_produced/06_dat_a.csv")
dat_q <- read.csv("Data_produced/06_dat_q.csv")
```

### b. Check variables
```{r}
unique(dat_a$Variable)
```

### c. Make data on broad format
```{r}
dat <- 
  dat_a %>% 
  spread(Variable, Value)
```


### d. Add NAO data
```{r}
df_nao <- read.table("https://climatedataguide.ucar.edu/sites/default/files/nao_station_djfm.txt", skip = 1, header = FALSE)
colnames(df_nao) <- c("Year", "Winter_NAO")

dat <- left_join(dat, df_nao)

```

### e. Wind data ??? Dag you mentioned you had files for Færder and Torungen?


### f. Checking missing years in dataset
* NB! TSM has been interpolated for 2012-2013, we decided not to interpolate the _Deep nutrients, because tehse are at the end of the time series
```{r}

str(dat)

check <-  dat %>%
  summarise_all(funs(sum(is.na(.))))

complete.cases(dat)
# there are missing values for four first years (1990-1994) and the last four years (2013-2016)
  #River TOC missing in 1990 and 1991, 
  #P measurements from hydro missing 1990
  #Plankton group data start in 1994
  #River_Si starts in 1995
  # DIN_deep, POC_deep, PON,_deep and POP_deep is missing from 2013-2016

# Suggestion: 
  #1. We cut 1990 and 1991 from the dataset (many missing variables)
  #2. we cut variables that have more than one year missing (have them separate below so we can test)

# NB! keep in mind that the deep POC and PON are the variables with highest upward trends in the MannKendal test. But TSM_deep is sampled over the whole time period, and has fairly high correlation with POC (0.63), also high correlations between deep and intermediate POC (0.75) and PON (0.74), same for DIN (0.72)
plot (dat$Hydro_TSM_Deep, dat$Hydro_POC_Deep)
plot (dat$Hydro_TSM_Deep, dat$Hydro_PON_Deep)
plot (dat$Hydro_POC_Deep, dat$Hydro_POC_Intermediate)
plot (dat$Hydro_DIN_Deep, dat$Hydro_DIN_Intermediate)
cor(dat$Hydro_TSM_Deep, dat$Hydro_POC_Deep, use = "complete.obs")
cor(dat$Hydro_PON_Deep, dat$Hydro_PON_Intermediate, use = "complete.obs")
cor(dat$Hydro_DIN_Deep, dat$Hydro_DIN_Intermediate, use = "complete.obs")

# filter the datasets according to suggestion above
cols <- c("Hydro_DIN_Deep", "Hydro_POC_Deep", "Hydro_PON_Deep", "Hydro_POP_Deep", "River_Si")

dat_sel <- dat %>% 
  filter(Year != 1990 & Year != 1991) %>% 
  dplyr::select(-starts_with('Plank')) %>% 
  dplyr::select(-one_of(cols))

```


# PART ONE - Hydrographic data

## 1. PCA of environmental variables
```{r environmental PCA, fig.height = 10, fig.width = 10}
#tail(dat)

# only complete cases now, se 1f above for selection
complete.cases(dat_sel)
dat.pca <- dat_sel[complete.cases(dat_sel), ]

##### on all vars in dat_sel
# ser på miljøvaribelrommet
var.pca <- rda(dat.pca[,-1], scale = TRUE)
envvar  <- data.frame(scores(var.pca, display = "species"))
#head(envvar)
years   <- data.frame(scores(var.pca, display = "sites"))
# make dataset for ggplot
pca_all <- bind_rows(envvar, years)
pca_all$obj <- c(row.names(envvar), dat.pca$Year)
pca_all$type <- c(rep("env", nrow(envvar)), rep("yr", nrow(years)))
# create ggplot
envvarplot_all <- pca_all %>% 
  ggplot(aes(PC1, PC2)) +
  geom_point(data = pca_all[pca_all$type=="yr", ], aes(color = as.numeric(as.character(obj)))) +
  labs(color = "Year") +
  geom_segment(data = pca_all[pca_all$type=="env", ], aes(x=0, xend=PC1*10, y=0, yend=PC2*10), 
                 linetype="dashed", 
                 arrow = arrow(length = unit(0.2, "cm"), type="closed"),
                 size = 0.1, col = "dark gray") +
  geom_text(data = pca_all[pca_all$type=="env", ], aes(x = PC1*10, y = PC2*10, label = obj),  color = "red", size = 3)

envvarplot_all

##### on all hydro vars
str (dat.pca)
hydro.pca <- dat.pca %>% 
  dplyr::select(starts_with("Yea"), starts_with("Hydro"))

# ser på miljøvaribelrommet
var.pca <- rda(hydro.pca[,-1], scale = TRUE)
envvar  <- data.frame(scores(var.pca, display = "species"))
#head(envvar)
years   <- data.frame(scores(var.pca, display = "sites"))
# make dataset for ggplot
pca_all <- bind_rows(envvar, years)
pca_all$obj <- c(row.names(envvar), hydro.pca$Year)
pca_all$type <- c(rep("env", nrow(envvar)), rep("yr", nrow(years)))
# create ggplot
envvarplot_hydro <- pca_all %>% 
  ggplot(aes(PC1, PC2)) +
  geom_point(data = pca_all[pca_all$type=="yr", ], aes(color = as.numeric(as.character(obj)))) +
  labs(color = "Year") +
  geom_segment(data = pca_all[pca_all$type=="env", ], aes(x=0, xend=PC1*10, y=0, yend=PC2*10), 
                 linetype="dashed", 
                 arrow = arrow(length = unit(0.2, "cm"), type="closed"),
                 size = 0.1, col = "dark gray") +
  geom_text(data = pca_all[pca_all$type=="env", ], aes(x = PC1*10, y = PC2*10, label = obj),  color = "red", size = 3)

  
envvarplot_hydro

##### on all river vars
str (dat.pca)
river.pca <- dat.pca %>% 
  dplyr::select(starts_with("Yea"), starts_with("River"))

# ser på miljøvaribelrommet
var.pca <- rda(river.pca[,-1], scale = TRUE)
envvar  <- data.frame(scores(var.pca, display = "species"))
#head(envvar)
years   <- data.frame(scores(var.pca, display = "sites"))
# make dataset for ggplot
pca_all <- bind_rows(envvar, years)
pca_all$obj <- c(row.names(envvar), river.pca$Year)
pca_all$type <- c(rep("env", nrow(envvar)), rep("yr", nrow(years)))
# create ggplot
envvarplot_river <- pca_all %>% 
  ggplot(aes(PC1, PC2)) +
  geom_point(data = pca_all[pca_all$type=="yr", ], aes(color = as.numeric(as.character(obj)))) +
  labs(color = "Year") +
  geom_segment(data = pca_all[pca_all$type=="env", ], aes(x=0, xend=PC1*2, y=0, yend=PC2*2), 
                 linetype="dashed", 
                 arrow = arrow(length = unit(0.2, "cm"), type="closed"),
                 size = 0.1, col = "dark gray") +
  geom_text(data = pca_all[pca_all$type=="env", ], aes(x = PC1*2, y = PC2*2, label = obj),  color = "red", size = 3)

  
envvarplot_river

```

## 2. Regression analyses



### Linear models
```{r}
### a. Surface salinity
full_model <- lm(Salinity_Surface ~ River_Local_Discharge + River_Distant_Discharge + Winter_NAO, data = dat)
# summary(full_model)
step_model <- stepAIC(full_model, direction = "both", trace = FALSE)
summary(step_model)

### b. Surface Total N
full_model <- lm(TotN_Surface ~ River_Local_Discharge + River_Distant_Discharge + Winter_NAO +
                 River_Local_TotN + River_Distant_TotN, data = dat)
# summary(full_model)
step_model <- stepAIC(full_model, direction = "both", trace = FALSE)
summary(step_model)

### c. Surface Total P
full_model <- lm(TotP_Surface ~ River_Local_Discharge + River_Distant_Discharge + Winter_NAO +
                 River_Local_TotP + River_Distant_TotP, data = dat)
# summary(full_model)
step_model <- stepAIC(full_model, direction = "both", trace = FALSE)
summary(step_model)

### d. Surface TSM
full_model <- lm(TSM_Surface ~ River_Local_Discharge + River_Distant_Discharge + Winter_NAO +
                 River_Local_TOC + River_Distant_TOC, data = dat)
# summary(full_model)
step_model <- stepAIC(full_model, direction = "both", trace = FALSE)
summary(step_model)
par(mfrow = c(2,2), mar = c(4,5,2,1))
plot(step_model)
visreg(step_model, points = list(cex = 1))

### e. Siktedyp
full_model <- lm(Secchi_Surface ~ River_Local_Discharge + River_Distant_Discharge + Winter_NAO +
                 River_Local_TotN + River_Distant_TotN + River_Local_TotP + River_Distant_TotP + 
                 River_Local_TOC + River_Distant_TOC, data = dat)
# summary(full_model)
step_model <- stepAIC(full_model, direction = "both", trace = FALSE)
summary(step_model)
visreg(step_model, points = list(cex = 1))

## 3. Hydrographic data, intermediate

### a. Intermediate salinity
full_model <- lm(Salinity_Intermediate ~ River_Local_Discharge + River_Distant_Discharge + Winter_NAO, data = dat)
# summary(full_model)
step_model <- stepAIC(full_model, direction = "both", trace = FALSE)
summary(step_model)

### b. Intermediate Total N
full_model <- lm(TotN_Intermediate ~ River_Local_Discharge + River_Distant_Discharge + Winter_NAO +
                 River_Local_TotN + River_Distant_TotN, data = dat)
# summary(full_model)
step_model <- stepAIC(full_model, direction = "both", trace = FALSE)
summary(step_model)

### c. Intermediate Total P
full_model <- lm(TotP_Intermediate ~ River_Local_Discharge + River_Distant_Discharge + Winter_NAO +
                 River_Local_TotP + River_Distant_TotP, data = dat)
# summary(full_model)
step_model <- stepAIC(full_model, direction = "both", trace = FALSE)
summary(step_model)

### d. Intermediate TSM
full_model <- lm(TSM_Intermediate ~ River_Local_Discharge + River_Distant_Discharge + Winter_NAO +
                 River_Local_TOC + River_Distant_TOC, data = dat)
# summary(full_model)
step_model <- stepAIC(full_model, direction = "both", trace = FALSE)
summary(step_model)
# par(mfrow = c(2,2), mar = c(4,5,2,1))
# plot(step_model)
par(mfrow = c(2,2), mar = c(4,5,2,1))
visreg(step_model, points = list(cex = 1))

## 4. Hydrographic data, deep

### a. Deep salinity
full_model <- lm(Salinity_Deep ~ River_Local_Discharge + River_Distant_Discharge + Winter_NAO, data = dat)
# summary(full_model)
step_model <- stepAIC(full_model, direction = "both", trace = FALSE)
summary(step_model)

### b. Deep Total N
full_model <- lm(TotN_Deep ~ River_Local_Discharge + River_Distant_Discharge + Winter_NAO +
                 River_Local_TotN + River_Distant_TotN, data = dat)
# summary(full_model)
step_model <- stepAIC(full_model, direction = "both", trace = FALSE)
summary(step_model)

### c. Deep Total P
full_model <- lm(TotP_Deep ~ River_Local_Discharge + River_Distant_Discharge + Winter_NAO +
                 River_Local_TotP + River_Distant_TotP, data = dat)
# summary(full_model)
step_model <- stepAIC(full_model, direction = "both", trace = FALSE)
summary(step_model)
par(mfrow = c(2,2), mar = c(4,5,2,1))
visreg(step_model, points = list(cex = 1))

### d. Deep TSM
full_model <- lm(TSM_Deep ~ River_Local_Discharge + River_Distant_Discharge + Winter_NAO +
                 River_Local_TOC + River_Distant_TOC, data = dat)
# summary(full_model)
step_model <- stepAIC(full_model, direction = "both", trace = FALSE)
summary(step_model)
# par(mfrow = c(2,2), mar = c(4,5,2,1))
# plot(step_model)
visreg(step_model, points = list(cex = 1))
```

## 5. Plankton
### a. Kiselalger
```{r}
full_model <- lm(Kiselalger_med ~ River_Local_Discharge + River_Distant_Discharge + Winter_NAO +
                 River_Local_TotN + River_Distant_TotN + River_Local_TotP + River_Distant_TotP + 
                 River_Local_TOC + River_Distant_TOC, data = dat)
# summary(full_model)
step_model <- stepAIC(full_model, direction = "both", trace = FALSE)
summary(step_model)
par(mfrow = c(2,2), mar = c(4,5,2,1))
visreg(step_model, points = list(cex = 1))



### b. Dinoflagellater
```{r}
full_model <- lm(Dinoflagellater_med ~ River_Local_Discharge + River_Distant_Discharge + Winter_NAO +
                 River_Local_TotN + River_Distant_TotN + River_Local_TotP + River_Distant_TotP + 
                 River_Local_TOC + River_Distant_TOC, data = dat)
# summary(full_model)
step_model <- stepAIC(full_model, direction = "both", trace = FALSE)
summary(step_model)
par(mfrow = c(2,2), mar = c(4,5,2,1))
visreg(step_model, points = list(cex = 1))

```

### c. Flagellater
```{r}
full_model <- lm(Dinoflagellater_med ~ River_Local_Discharge + River_Distant_Discharge + Winter_NAO +
                 River_Local_TotN + River_Distant_TotN + River_Local_TotP + River_Distant_TotP + 
                 River_Local_TOC + River_Distant_TOC, data = dat)
# summary(full_model)
step_model <- stepAIC(full_model, direction = "both", trace = FALSE)
summary(step_model)
par(mfrow = c(2,2), mar = c(4,5,2,1))
visreg(step_model, points = list(cex = 1))

```

```


# PART TWO - Biologigal data

We're trying to find meaningful ways to describe the relationship between biological paramteres and environmental variables, and connect this to changes that have happened over time.

** Tentative approach: **

* Plot environmental vectors on ordination plots
* Use this to generate hypotheses about causal links
* Test by using models 
  + Probably different models for every system
  + Mixed effects for hard bottom (station as random)
  + GLM for soft bottom (one model per station for DCA at least) and plankton
* Probably makes sense to use the same set of explanatory variables in model selection for every response witin a system
* Should perhaps add some variables as factors.

## Hard bottom
### Environmental variables in relation to ordination
```{r fitting environmental variables to ordination}
dim(dat)

# retrieving HB parameters
HBdf <- readr::read_csv2("Datasett/Hardbunn_KOPI/HBanalysesett.csv")
names(HBdf)

# coupling the two using dplyr, by Year
HBdat <- # joining and removing phyto from the nix
  HBdf %>% 
    filter(Site %in% c(407, 410)) %>% # filter on stations 407 and 410
    left_join(dat, by = "Year") %>% # join
    dplyr::select(-ends_with("_med")) # remove phytopl

DCAhb <- HBdat %>% 
  dplyr::select(DCA1:DCA4)

envHB <- HBdat %>% 
  dplyr::select(Chla_Deep:Winter_NAO)

envfitHB <- envfit(DCAhb, envHB, na.rm = TRUE, permutations = 999, choices = 1:4)
envfitHB # gives an indication of which parameters may be important for community responses along dca 1, 12 obs fjernes pga NA
str(envfitHB)

plot(DCAhb[1:2])
plot(envfitHB, p.max = 0.01, add = TRUE) # plotting parameters with p-value < 0.01 in relation to ordination. Can make this into a ggplot if needed

HBpar <- names(which(envfitHB$vectors$pvals < 0.01)) # names of parameters with p-value < 0.01
# Must remove phytoplankton
```

### DCA - possible causal links
```{r fitting model for DCA}

# For use as random factor
HBdat$Site <- as.character(HBdat$Site)

# Standarizing
# https://cran.r-project.org/web/packages/standardize/vignettes/using-standardize.html 

HBfullformula <- as.formula(paste("DCA1 ~ ", paste(HBpar, collapse = " + "), "+ (1|Site)", sep = ""))

sHB <- standardize(HBfullformula, HBdat)
HBlmer <- lmer(sHB$formula, sHB$data)

# Model selection
# https://cran.r-project.org/web/packages/lmerTest/lmerTest.pdf 

HBlmer_select <- lmerTest::step(HBlmer) # check envvar for missing values
HBlmer_select

# a bit worried about autocorrelation over time... not sure how much of an issue it is with this kind of data... Dag?

```

## Soft bottom
### Environmental variables in relation to ordination
```{r fitting environmental variables to ordination}
dim(dat)

# retrieving BB ordinations
BB05df <- readr::read_csv2("Data_produced/BBordSites_05.csv")
names(BB05df)
BB35df <- readr::read_csv2("Data_produced/BBordSites_35.csv")
names(BB35df)

# coupling the two using dplyr, by Year
BB05_dat <- # joining and removing phyto from the mix
  BB05df %>% 
    left_join(dat, by = "Year") %>% # join
    dplyr::select(-ends_with("_med")) # remove phytopl

DCA05 <- BB05_dat %>% 
  dplyr::select(DCA1:DCA4)

DCA35 <- BB35df %>% 
  dplyr::select(DCA1:DCA4)

envBB <- BB05_dat %>% 
  dplyr::select(Chla_Deep:Winter_NAO)


BB35_dat <- bind_cols(BB35df[,-1], envBB)

envfitBB05 <- envfit(DCA05, envBB, na.rm = TRUE, permutations = 999, choices = 1:4)
envfitBB05 # gives an indication of which parameters may be important for community responses 
# 6 obs fjernes pga NA
envfitBB35 <- envfit(DCA35, envBB, na.rm = TRUE, permutations = 999, choices = 1:4)
envfitBB35 
# 6 obs fjernes pga NA


plot(DCA05[1:2])
plot(envfitBB05, p.max = 0.05, add = TRUE)
plot(DCA35[1:2])
plot(envfitBB35, p.max = 0.05, add = TRUE)

# plotting parameters with p-value < 0.01 in relation to ordination. Can make this into a ggplot if needed

BB05par <- names(which(envfitBB05$vectors$pvals < 0.05)) # names of parameters with p-value < 0.05
BB35par <- names(which(envfitBB35$vectors$pvals < 0.05)) # names of parameters with p-value < 0.05

```

### DCA - possible causal links
```{r fitting model for DCA}

# Standarizing
# https://cran.r-project.org/web/packages/standardize/vignettes/using-standardize.html 

BB05_fullformula <- as.formula(paste("DCA1 ~ ", paste(BB05par, collapse = " + "), sep = ""))
BB35_fullformula <- as.formula(paste("DCA1 ~ ", paste(BB35par, collapse = " + "), sep = ""))

sBB05 <- standardize(BB05_fullformula, BB05_dat)
sBB35 <- standardize(BB35_fullformula, BB35_dat)

# Model and selection
BB05_lm <- lm(sBB05$formula, sBB05$data)
BB05_select <- stepAIC(BB05_lm, direction = "both") # check envvar for missing values - will this be fixed?
BB05_select

BB05_dat %>%
  summarise_all(funs(sum(is.na(.))))

BB35_lm <- lm(sBB35$formula, sBB35$data)
BB35_select <- stepAIC(BB35_lm, direction = "both") 
BB35_select

```

## Plankton
### Environmental variables in relation to ordination
```{r fitting environmental variables to ordination}
dim(dat)

# retrieving BB ordinations
P01df <- readr::read_csv2("Data_produced/P01ord.csv")
names(P01df)


# coupling the two using dplyr, by Year
P01_dat <- # joining and removing phyto from the mix
  P01df %>% 
    left_join(dat, by = "Year") %>% # join
    dplyr::select(-ends_with("_med")) # remove phytopl

DCA01 <- P01_dat %>% 
  dplyr::select(DCA1:DCA4)

envP01 <- P01_dat %>% 
  dplyr::select(Chla_Deep:Winter_NAO)

envfitP01 <- envfit(DCA01, envP01, na.rm = TRUE, permutations = 999, choices = 1:4)
envfitP01 # gives an indication of which parameters may be important for community responses 

plot(DCA01[1:2])
plot(envfitP01, p.max = 0.05, add = TRUE)

# plotting parameters with p-value < 0.01 in relation to ordination. Can make this into a ggplot if needed
# for plankton it looks like we might need to something about the ordination

P01par <- names(which(envfitP01$vectors$pvals < 0.05)) # names of parameters with p-value < 0.05

```

### DCA - possible causal links
```{r fitting model for DCA}

# Standarizing
# https://cran.r-project.org/web/packages/standardize/vignettes/using-standardize.html 

P01_fullformula <- as.formula(paste("DCA1 ~ ", paste(P01par, collapse = " + "), sep = ""))

sP01 <- standardize(P01_fullformula, P01_dat)

# Model and selection
P01_lm <- lm(sP01$formula, sP01$data)
P01_select <- stepAIC(P01_lm, direction = "both") # check envvar for missing values - will this be fixed?
P01_select

P01_dat %>%
  summarise_all(funs(sum(is.na(.))))

summary(P01_select)
```
